Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-08-04 latex/manuscript-main_text.tex
Merge pull request #81 from greenelab/rev2 [ci skip]
This build is based on
https://github.com/greenelab/phenoplier_manuscript/commit/01398b0990ea0f8bbdbb98e79c0585e2d905a51f.

This commit was created by the following CI build and job:
https://github.com/greenelab/phenoplier_manuscript/commit/01398b0990ea0f8bbdbb98e79c0585e2d905a51f/checks
https://github.com/greenelab/phenoplier_manuscript/actions/runs/5765234476

Each LV or gene module represents a group of weighted genes expressed together in the same tissues and cell types as a functional unit.
Since LVs might represent a functional set of genes regulated by the same transcriptional program\textsuperscript{\protect\hyperlink{ref-1FVd2WW6G}{48},\protect\hyperlink{ref-DN7TyZzb}{49}}, we conjecture that the projection of TWAS and pharmacologic perturbations data into this latent space could provide a better mechanistic understanding.
For this projection of different data modalities into the same space, PhenoPLIER converts gene associations to an LV score: all genes' standardized effect sizes for a trait (from TWAS) or differential expression values for a drug (from pharmacologic perturbation data) are multiplied by the LV genes' weights and summed, producing a single value.
Instead of looking at individual genes, this process links different traits and drugs to functionally-related groups of genes or LVs.
PhenoPLIER uses LVs annotations about the specific conditions where the group of genes is expressed, such as cell types and tissues, even at specific developmental stages, disease stages or under distinct stimuli.
Although this is not strictly necessary for PhenoPLIER to work, these annotations can dramatically improve the interpretability of results.
MultiPLIER's models provide this information by linking LVs to samples, which may be annotated for experimental conditions (represented by matrix \(\mathbf{B}\) at the top of Figure \ref{fig:entire_process}b) in which genes in an LV are expressed.
An example of this is shown in Figure \ref{fig:entire_process}c.
In the original MultiPLIER study, the authors reported that one of the latent variables, identified as LV603, was associated with a known neutrophil pathway and highly correlated with neutrophil count estimates from whole blood RNA-seq profiles\textsuperscript{\protect\hyperlink{ref-18TSqd1tG}{50}}.
We analyzed LV603 using PhenoPLIER and found that
1) neutrophil counts and other white blood cell traits were ranked among the top 10 traits out of 4,091 (Figure \ref{fig:entire_process}c, bottom), and basophils count and percentage were significantly associated with this LV when using our regression method (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv603}),
and 2) LV603's genes were expressed in highly relevant cell types (Figure \ref{fig:entire_process}c, top).
These initial results suggested that groups of functionally related and co-expressed genes tend to correspond to groups of trait-associated genes, and the approach can link transcriptional mechanisms from large and diverse dataset collections to complex traits.

@@

Next, we analyzed all 987 LVs using Fast Gene Set Enrichment Analysis (FGSEA)\textsuperscript{\protect\hyperlink{ref-Z8WXLD67}{51}}, and found 15 LVs nominally enriched (unadjusted \emph{P} \textless{} 0.01) with these lipid-altering gene-sets (Supplementary Tables \ref{tbl:sup:lipids_crispr:modules_enriched_increase} and \ref{tbl:sup:lipids_crispr:modules_enriched_decrease}).
Among those with reliable sample metadata, LV246, the top LV associated with the lipids-increasing gene-set, contained genes mainly co-expressed in adipose tissue (Figure \ref{fig:lv246}a), which plays a key role in coordinating and regulating lipid metabolism.
Using our regression framework across all traits in PhenomeXcan, we found that gene weights for this LV were predictive of gene associations for plasma lipids, high cholesterol, and Alzheimer's disease (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv246}, FDR \textless{} 1e-23).
These lipids-related associations also replicated across the 309 traits in eMERGE (Supplementary Table \ref{tbl:sup:emerge_assocs:lv246}), where LV246 was significantly associated with hypercholesterolemia (phecode: 272.11, FDR \textless{} 4e-9), hyperlipidemia (phecode: 272.1, FDR \textless{} 4e-7) and disorders of lipoid metabolism (phecode: 272, FDR \textless{} 4e-7).

@@

We examined a specific drug-disease pair to determine whether the LVs driving the prediction were biologically plausible.
Nicotinic acid (niacin) is a B vitamin widely used clinically to treat lipid disorders, although there is controversy on its clinical utility in preventing cardiovascular disease\textsuperscript{\protect\hyperlink{ref-bRPc66OD}{58}--\protect\hyperlink{ref-ZGvG75Bj}{60}}.
Niacin exerts its effects on multiple tissues, although its mechanisms are not well understood\textsuperscript{\protect\hyperlink{ref-LVihFr3g}{61}--\protect\hyperlink{ref-13meq3Hgt}{64}}.
This compound can increase high-density lipoprotein (HDL) by inhibiting an HDL catabolism receptor in the liver.
Niacin also inhibits diacylglycerol acyltransferase--2 (DGAT2), which decreases the production of low-density lipoproteins (LDL) either by modulating triglyceride synthesis in hepatocytes or by inhibiting adipocyte triglyceride lipolysis\textsuperscript{\protect\hyperlink{ref-LVihFr3g}{61}}.
Niacin was one of the drugs in the gold standard set indicated for atherosclerosis (AT) and coronary artery disease (CAD).
We observed that this compound was predicted by the gene-based and LV-based approach as a medical indication for coronary artery disease (CAD), with scores above the mean (0.51 and 0.96, respectively).
For AT, the LV-based approach predicted niacin as a therapeutic drug with a score of 0.52, whereas the gene-based method assigned a negative score of -0.01 (below the mean).
Since LVs represent interpretable features associated with specific cell types, we analyzed which LVs positively contributed to these predictions (i.e., with an opposite direction between niacin and the disease).
Notably, LV246 (Figure \ref{fig:lv246}), expressed in adipose tissue and liver and associated with plasma lipids and high cholesterol (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv246}), was the 16th most important module in the prediction of niacin as a therapeutic drug for AT.
Besides the gold standard set, LV246 was among the top modules for other cardiovascular diseases, such as ischaemic heart disease (wide definition, 15th module) and high cholesterol (7th module).

The analysis of other top niacin-contributing LVs across different cardiovascular diseases revealed additional mechanisms of action.
For example, \emph{GPR109A/HCAR2} encodes a G protein-coupled high-affinity niacin receptor in adipocytes and immune cells, including monocytes, macrophages, neutrophils and dendritic cells\textsuperscript{\protect\hyperlink{ref-izbPw2kc}{65},\protect\hyperlink{ref-wI0IjT3i}{66}}.
It was initially thought that the antiatherogenic effects of niacin were solely due to the inhibition of lipolysis in adipose tissue.
However, it has been shown that nicotinic acid can reduce atherosclerosis progression independently of its antidyslipidemic activity by activating \emph{GPR109A} in immune cells\textsuperscript{\protect\hyperlink{ref-1Bz0jRHYo}{67}}, thus boosting anti-inflammatory processes\textsuperscript{\protect\hyperlink{ref-1DblG8swn}{68}}.
In addition, flushing, a common adverse effect of niacin, is also produced by the activation of GPR109A in Langerhans cells (macrophages of the skin).
This alternative mechanism for niacin could have been hypothesized by examining the cell types where the top-contributing modules are expressed:
for instance, LV116 and LV931 (Figure \ref{fig:lv116:cell_types}, Supplementary Figure \ref{fig:sup:lv931}, and Supplementary Tables \ref{tbl:sup:multiplier_pathways:lv116} and \ref{tbl:sup:multiplier_pathways:lv931}) were the top two modules for AT, with a strong signature in monocytes, macrophages, neutrophils, dendritic cells, among others.
In Figure \ref{fig:lv116:cell_types}, it can be seen that LV116's genes are expressed as an immune response when these cell types are under different stimuli, such as diarrhea caused by different pathogens\textsuperscript{\protect\hyperlink{ref-RliFvowC}{69}}, samples from multiple sclerosis or systemic lupus erythematosus\textsuperscript{\protect\hyperlink{ref-D5XBhzim}{70},\protect\hyperlink{ref-EnoqU4ga}{71}}, or infected with different viruses (such as herpes simplex\textsuperscript{\protect\hyperlink{ref-11eausmiy}{72}}, West Nile virus\textsuperscript{\protect\hyperlink{ref-mtMYROCN}{73}}, \emph{Salmonella typhimurium}\textsuperscript{\protect\hyperlink{ref-1kgcHkGm}{74}}, among others).
These three LVs (LV246, LV116 and LV931) were among the top 20 modules contributing to the niacin prediction across different cardiovascular traits (Table \ref{tbl:niacin:cardio:top_lvs}).

Beyond cardiovascular traits, there are other potentially interesting LVs that could extend our understanding of the mechanisms of niacin.
For example, LV66, one of the top LVs affected by niacin (Supplementary Figure \ref{fig:sup:lv66}), was mainly expressed in ovarian granulosa cells.
This compound has been very recently considered a potential therapeutic for ovarian diseases\textsuperscript{\protect\hyperlink{ref-kLRErKXz}{75},\protect\hyperlink{ref-TovvsrDr}{76}}, as it was found to promote follicle growth and inhibit granulosa cell apoptosis in animal models.

@@

We used the projection of gene-trait associations into the latent space to find groups of clusters linked by the same transcriptional processes.
Since individual clustering algorithms have different biases (i.e., assumptions about the data structure), we designed a consensus clustering framework that combines solutions or partitions of traits generated by different methods (\protect\hyperlink{sec:methods:clustering}{Methods}).
Consensus or ensemble approaches have been recommended to avoid several pitfalls when performing cluster analysis on biological data\textsuperscript{\protect\hyperlink{ref-14dCeRkua}{77}}.
Since diversity in the ensemble is crucial for these methods, we generated different data versions which were processed using different methods with varying sets of parameters (Figure \ref{fig:clustering:design}a).
Then, a consensus function combines the ensemble into a consolidated solution, which has been shown to outperform any individual member of the ensemble\textsuperscript{\protect\hyperlink{ref-uw3AnEgA}{78},\protect\hyperlink{ref-cuROQDFa}{79}}.
Our clustering pipeline generated 15 final consensus clustering solutions (Supplementary Figure \ref{fig:sup:clustering:agreement}).
The number of clusters of these partitions (between 5 to 29) was learned from the data by selecting the partitions with the largest agreement with the ensemble\textsuperscript{\protect\hyperlink{ref-uw3AnEgA}{78}}.
Instead of selecting one of these final solutions with a specific number of clusters, we used a clustering tree\textsuperscript{\protect\hyperlink{ref-xhtEAzx6}{80}} (Figure \ref{fig:clustering:tree}) to examine stable groups of traits across multiple resolutions.
To understand which latent variables differentiated the group of traits, we trained a decision tree classifier on the input data \(\hat{\mathbf{M}}\) using the clusters found as labels (Figure \ref{fig:clustering:design}b, see \protect\hyperlink{sec:methods:clustering}{Methods}).

@@

Next, we analyzed which LVs were driving these clusters of traits.
For this, we trained decision tree classifiers on the input data using each cluster at \(k\)=29 (bottom of Figure \ref{fig:clustering:tree}) as labels (see \protect\hyperlink{sec:methods:clustering}{Methods}).
This procedure yielded the top LVs that were most discriminative for each cluster.
Several of these LVs were well-aligned to existing pathways (Figure \ref{fig:clustering:heatmap}), whereas others were not aligned to prior knowledge but still expressed in relevant tissues (Supplementary Figure \ref{fig:sup:clustering:novel:heatmap}).
In Figure \ref{fig:clustering:heatmap}, it can be seen that some LVs were highly specific to certain traits, while others were associated with a wide range of different phenotypes, thus potentially involved in more general biological functions.
We used our regression framework to determine whether these LVs were significantly associated with different traits.
For example, LVs such as LV928 and LV30, which were well-aligned to early progenitors of the erythrocytes lineage\textsuperscript{\protect\hyperlink{ref-Zk82GvJV}{82}} (Supplementary Tables \ref{tbl:sup:multiplier_pathways:lv928} and \ref{tbl:sup:multiplier_pathways:lv30}), were predominantly expressed in early differentiation stages of erythropoiesis (Supplementary Figures \ref{fig:sup:lv928} and \ref{fig:sup:lv30}) and strongly associated with different assays on red blood cells (FDR \textless{} 0.05; Supplementary Tables \ref{tbl:sup:phenomexcan_assocs:lv928}, \ref{tbl:sup:emerge_assocs:lv928}, and \ref{tbl:sup:emerge_assocs:lv30}).
In contrast, other LVs were highly specific, such as LV730, which is expressed in thrombocytes from different cancer samples (Supplementary Figures \ref{fig:sup:lv730} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv730}), and strongly associated with hematological assays on platelets (FDR \textless{} 0.05, Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv730});
or LV598, whose genes were expressed in corneal endothelial cells (Supplementary Figures \ref{fig:sup:lv598} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv598}) and associated with keratometry measurements (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv598}).

The sub-branches of autoimmune and cardiovascular diseases merged together at \(k=10\) (middle of Figure \ref{fig:clustering:tree}), so we expected to find LVs that specifically affect one or both of these types of diseases.
For example, LV57, expressed in T cells (Supplementary Figure \ref{fig:sup:lv57} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv57}), was the most strongly associated gene module with autoimmune disorders in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv57}), with significant associations with hypothyroidism that were replicated in eMERGE (Supplementary Table \ref{tbl:sup:emerge_assocs:lv57}).
However, this LV was also strongly associated with deep venous thrombosis in both PhenomeXcan and eMERGE.
On the other hand, LV844 was more autoimmune-specific, with associations to polymyalgia rheumatica, type 1 diabetes, rheumatoid arthritis, and celiac disease in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv844}).
However, these did not replicate in eMERGE.
This LV was expressed in a wide range of cell types, including blood, breast organoids, myeloma cells, lung fibroblasts, and different cell types from the brain (Supplementary Figure \ref{fig:sup:lv844} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv844}).

The cardiovascular sub-branch had 129 significant LV-trait associations in PhenomeXcan and 23 in eMERGE.
LV136, aligned with known collagen formation and muscle contraction pathways (Supplementary Table \ref{tbl:sup:multiplier_pathways:lv136}), was associated with coronary artery disease and keratometry measurements in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv136}).
In eMERGE, this LV was associated with coronary atherosclerosis (phecode: 411.4) (Supplementary Table \ref{tbl:sup:emerge_assocs:lv136}).
LV136 was expressed in a wide range of cell types, including fibroblasts, mesenchymal stem cells, osteoblasts, pancreatic stellate cells, cardiomyocytes, and adipocytes (Supplementary Figure \ref{fig:sup:lv136}).
Within the cardiovascular sub-branch, we found neuropsychiatric and neurodevelopmental disorders such as Alzheimer's disease, schizophrenia, and attention deficit hyperactivity disorder (ADHD).
These disorders were previously linked to the cardiovascular system\textsuperscript{\protect\hyperlink{ref-10qjLoufR}{83}--\protect\hyperlink{ref-13t4TuFeJ}{86}} and share several risk factors, including hypertension, high cholesterol, obesity, smoking, among others\textsuperscript{\protect\hyperlink{ref-D83Aqhga}{87},\protect\hyperlink{ref-9BGyO071}{88}}.
However, our results grouped these diseases by potentially shared transcriptional processes expressed in specific tissues/cell types.
Alzheimer's disease (not present in eMERGE), for instance, was significantly associated with LV21 in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv21}).
LV21, a gene module not aligned to prior pathways, was strongly expressed in a variety of soft tissue sarcomas, monocytes/macrophages (including microglia from cortex samples), and aortic valves (Supplementary Figure \ref{fig:sup:lv21} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv21}).
This LV was also strongly associated with lipids and high cholesterol in PhenomeXcan and hyperlipidemia (phecode: 272.1) in eMERGE (Supplementary Table \ref{tbl:sup:emerge_assocs:lv21}).
As discussed previously, macrophages play a key role in the reverse cholesterol transport and thus atherogenesis\textsuperscript{\protect\hyperlink{ref-idlBgtFz}{89}}, and lipid metabolism in microglia has been recently identified as an important factor in the development of neurodegenerative diseases\textsuperscript{\protect\hyperlink{ref-18I4ish9s}{90}}.

@@

Note that, since we used the MultiXcan regression model (Equation (\ref{eq:multixcan})), \(\mathbf{R}\) is only an approximation of gene correlations in S-MultiXcan.
As explained before, S-MultiXcan approximates the joint regression parameters in MultiXcan using the marginal regression estimates from S-PrediXcan in (\ref{eq:spredixcan}) with some simplifying assumptions and different genotype covariance matrices.
This complicates the derivation of an S-MultiXcan-specific solution to compute \(\mathbf{R}\).
To account for this, we used a submatrix \(\mathbf{R}_{\ell}\) corresponding to genes that are part of LV \(\ell\) only (top 1\% of genes) instead of the entire matrix \(\mathbf{R}\).
This simplification is conservative since correlations are accounted for top genes only.
Our simulations (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) show that the model is approximately well-calibrated and can correct for LVs with adjacent and highly correlated genes at the top (e.g., Supplementary Figure \ref{fig:reg:nulls:qqplot:lv234}).
The simulation also identified 127 LVs in which the model was not well-calibrated (e.g., Supplementary Figure \ref{fig:reg:nulls:qqplot:lv914}).
As this can be attributed to limitations in accurately computing a gene correlation matrix, we excluded these LVs from our main analyses.

@@

Finally, we used spectral clustering on \(\mathbf{D}\) to derive the final consensus partitions.
\(\mathbf{D}\) was first transformed into a similarity matrix by applying an RBF kernel \(\mathrm{exp}(-\gamma \mathbf{D}^2)\) using four different values for \(\gamma\) that we empirically determined to work best.
Therefore, for each \(k\) between 2 and 60, we derived four consensus partitions and selected the one that maximized Equation (\ref{eq:consensus:obj_func}).
We further filtered this set of 59 solutions to keep only those with an ensemble agreement larger than the 75th percentile (Supplementary Figure \ref{fig:sup:clustering:agreement}), leaving a total of 15 final consensus partitions shown in Figure \ref{fig:clustering:tree}.

>>>>

Each LV or gene module represents a group of weighted genes expressed together in the same tissues and cell types as a functional unit.
Since LVs might represent a functional set of genes regulated by the same transcriptional program\textsuperscript{\protect\hyperlink{ref-1FVd2WW6G}{48},\protect\hyperlink{ref-DN7TyZzb}{49}}, we conjecture that the projection of TWAS and pharmacologic perturbations data into this latent space could provide a better mechanistic understanding.
For this projection of different data modalities into the same space, PhenoPLIER converts gene associations to an LV score: all genes' standardized effect sizes for a trait (from TWAS) or differential expression values for a drug (from pharmacologic perturbation data) are multiplied by the LV genes' weights and summed, producing a single value.
Instead of looking at individual genes, this process links different traits and drugs to functionally-related groups of genes or LVs.
PhenoPLIER uses LVs annotations about the specific conditions where the group of genes is expressed, such as cell types and tissues, even at specific developmental stages, disease stages or under distinct stimuli.
Although this is not strictly necessary for PhenoPLIER to work, these annotations can dramatically improve the interpretability of results.
MultiPLIER's models provide this information by linking LVs to samples, which may be annotated for experimental conditions (represented by matrix \(\mathbf{B}\) at the top of Figure \ref{fig:entire_process}b) in which genes in an LV are expressed.
An example of this is shown in Figure \ref{fig:entire_process}c.
In the original MultiPLIER study, the authors reported that one of the latent variables, identified as LV603, was associated with a known neutrophil pathway and highly correlated with neutrophil count estimates from whole blood RNA-seq profiles\textsuperscript{\protect\hyperlink{ref-18TSqd1tG}{50}}.
We analyzed LV603 using PhenoPLIER and found that
1) neutrophil counts and other white blood cell traits were ranked among the top 10 traits out of 4,091 (Figure \ref{fig:entire_process}c, bottom), and basophils count and percentage were significantly associated with this LV when using our regression method (Supplementary Table 4),
and 2) LV603's genes were expressed in highly relevant cell types (Figure \ref{fig:entire_process}c, top).
These initial results suggested that groups of functionally related and co-expressed genes tend to correspond to groups of trait-associated genes, and the approach can link transcriptional mechanisms from large and diverse dataset collections to complex traits.

@@

Next, we analyzed all 987 LVs using Fast Gene Set Enrichment Analysis (FGSEA)\textsuperscript{\protect\hyperlink{ref-Z8WXLD67}{51}}, and found 15 LVs nominally enriched (unadjusted \emph{P} \textless{} 0.01) with these lipid-altering gene-sets (Supplementary Tables 1 and 2).
Among those with reliable sample metadata, LV246, the top LV associated with the lipids-increasing gene-set, contained genes mainly co-expressed in adipose tissue (Figure \ref{fig:lv246}a), which plays a key role in coordinating and regulating lipid metabolism.
Using our regression framework across all traits in PhenomeXcan, we found that gene weights for this LV were predictive of gene associations for plasma lipids, high cholesterol, and Alzheimer's disease (Supplementary Table 7, FDR \textless{} 1e-23).
These lipids-related associations also replicated across the 309 traits in eMERGE (Supplementary Table 8), where LV246 was significantly associated with hypercholesterolemia (phecode: 272.11, FDR \textless{} 4e-9), hyperlipidemia (phecode: 272.1, FDR \textless{} 4e-7) and disorders of lipoid metabolism (phecode: 272, FDR \textless{} 4e-7).

@@

We examined a specific drug-disease pair to determine whether the LVs driving the prediction were biologically plausible.
Nicotinic acid (niacin) is a B vitamin widely used clinically to treat lipid disorders, although there is controversy on its clinical utility in preventing cardiovascular disease\textsuperscript{\protect\hyperlink{ref-bRPc66OD}{58}--\protect\hyperlink{ref-ZGvG75Bj}{60}}.
Niacin exerts its effects on multiple tissues, although its mechanisms are not well understood\textsuperscript{\protect\hyperlink{ref-LVihFr3g}{61}--\protect\hyperlink{ref-13meq3Hgt}{64}}.
This compound can increase high-density lipoprotein (HDL) by inhibiting an HDL catabolism receptor in the liver.
Niacin also inhibits diacylglycerol acyltransferase--2 (DGAT2), which decreases the production of low-density lipoproteins (LDL) either by modulating triglyceride synthesis in hepatocytes or by inhibiting adipocyte triglyceride lipolysis\textsuperscript{\protect\hyperlink{ref-LVihFr3g}{61}}.
Niacin was one of the drugs in the gold standard set indicated for atherosclerosis (AT) and coronary artery disease (CAD).
We observed that this compound was predicted by the gene-based and LV-based approach as a medical indication for coronary artery disease (CAD), with scores above the mean (0.51 and 0.96, respectively).
For AT, the LV-based approach predicted niacin as a therapeutic drug with a score of 0.52, whereas the gene-based method assigned a negative score of -0.01 (below the mean).
Since LVs represent interpretable features associated with specific cell types, we analyzed which LVs positively contributed to these predictions (i.e., with an opposite direction between niacin and the disease).
Notably, LV246 (Figure \ref{fig:lv246}), expressed in adipose tissue and liver and associated with plasma lipids and high cholesterol (Supplementary Table 7), was the 16th most important module in the prediction of niacin as a therapeutic drug for AT.
Besides the gold standard set, LV246 was among the top modules for other cardiovascular diseases, such as ischaemic heart disease (wide definition, 15th module) and high cholesterol (7th module).

The analysis of other top niacin-contributing LVs across different cardiovascular diseases revealed additional mechanisms of action.
For example, \emph{GPR109A/HCAR2} encodes a G protein-coupled high-affinity niacin receptor in adipocytes and immune cells, including monocytes, macrophages, neutrophils and dendritic cells\textsuperscript{\protect\hyperlink{ref-izbPw2kc}{65},\protect\hyperlink{ref-wI0IjT3i}{66}}.
It was initially thought that the antiatherogenic effects of niacin were solely due to the inhibition of lipolysis in adipose tissue.
However, it has been shown that nicotinic acid can reduce atherosclerosis progression independently of its antidyslipidemic activity by activating \emph{GPR109A} in immune cells\textsuperscript{\protect\hyperlink{ref-1Bz0jRHYo}{67}}, thus boosting anti-inflammatory processes\textsuperscript{\protect\hyperlink{ref-1DblG8swn}{68}}.
In addition, flushing, a common adverse effect of niacin, is also produced by the activation of GPR109A in Langerhans cells (macrophages of the skin).
This alternative mechanism for niacin could have been hypothesized by examining the cell types where the top-contributing modules are expressed:
for instance, LV116 and LV931 (Figure \ref{fig:lv116:cell_types}, Supplementary Figure 15, and Supplementary Tables 9 and 10) were the top two modules for AT, with a strong signature in monocytes, macrophages, neutrophils, dendritic cells, among others.
In Figure \ref{fig:lv116:cell_types}, it can be seen that LV116's genes are expressed as an immune response when these cell types are under different stimuli, such as diarrhea caused by different pathogens\textsuperscript{\protect\hyperlink{ref-RliFvowC}{69}}, samples from multiple sclerosis or systemic lupus erythematosus\textsuperscript{\protect\hyperlink{ref-D5XBhzim}{70},\protect\hyperlink{ref-EnoqU4ga}{71}}, or infected with different viruses (such as herpes simplex\textsuperscript{\protect\hyperlink{ref-11eausmiy}{72}}, West Nile virus\textsuperscript{\protect\hyperlink{ref-mtMYROCN}{73}}, \emph{Salmonella typhimurium}\textsuperscript{\protect\hyperlink{ref-1kgcHkGm}{74}}, among others).
These three LVs (LV246, LV116 and LV931) were among the top 20 modules contributing to the niacin prediction across different cardiovascular traits (Table \ref{tbl:niacin:cardio:top_lvs}).

Beyond cardiovascular traits, there are other potentially interesting LVs that could extend our understanding of the mechanisms of niacin.
For example, LV66, one of the top LVs affected by niacin (Supplementary Figure 16), was mainly expressed in ovarian granulosa cells.
This compound has been very recently considered a potential therapeutic for ovarian diseases\textsuperscript{\protect\hyperlink{ref-kLRErKXz}{75},\protect\hyperlink{ref-TovvsrDr}{76}}, as it was found to promote follicle growth and inhibit granulosa cell apoptosis in animal models.

@@

We used the projection of gene-trait associations into the latent space to find groups of clusters linked by the same transcriptional processes.
Since individual clustering algorithms have different biases (i.e., assumptions about the data structure), we designed a consensus clustering framework that combines solutions or partitions of traits generated by different methods (\protect\hyperlink{sec:methods:clustering}{Methods}).
Consensus or ensemble approaches have been recommended to avoid several pitfalls when performing cluster analysis on biological data\textsuperscript{\protect\hyperlink{ref-14dCeRkua}{77}}.
Since diversity in the ensemble is crucial for these methods, we generated different data versions which were processed using different methods with varying sets of parameters (Figure \ref{fig:clustering:design}a).
Then, a consensus function combines the ensemble into a consolidated solution, which has been shown to outperform any individual member of the ensemble\textsuperscript{\protect\hyperlink{ref-uw3AnEgA}{78},\protect\hyperlink{ref-cuROQDFa}{79}}.
Our clustering pipeline generated 15 final consensus clustering solutions (Supplementary Figure 13).
The number of clusters of these partitions (between 5 to 29) was learned from the data by selecting the partitions with the largest agreement with the ensemble\textsuperscript{\protect\hyperlink{ref-uw3AnEgA}{78}}.
Instead of selecting one of these final solutions with a specific number of clusters, we used a clustering tree\textsuperscript{\protect\hyperlink{ref-xhtEAzx6}{80}} (Figure \ref{fig:clustering:tree}) to examine stable groups of traits across multiple resolutions.
To understand which latent variables differentiated the group of traits, we trained a decision tree classifier on the input data \(\hat{\mathbf{M}}\) using the clusters found as labels (Figure \ref{fig:clustering:design}b, see \protect\hyperlink{sec:methods:clustering}{Methods}).

@@

Next, we analyzed which LVs were driving these clusters of traits.
For this, we trained decision tree classifiers on the input data using each cluster at \(k\)=29 (bottom of Figure \ref{fig:clustering:tree}) as labels (see \protect\hyperlink{sec:methods:clustering}{Methods}).
This procedure yielded the top LVs that were most discriminative for each cluster.
Several of these LVs were well-aligned to existing pathways (Figure \ref{fig:clustering:heatmap}), whereas others were not aligned to prior knowledge but still expressed in relevant tissues (Supplementary Figure 14).
In Figure \ref{fig:clustering:heatmap}, it can be seen that some LVs were highly specific to certain traits, while others were associated with a wide range of different phenotypes, thus potentially involved in more general biological functions.
We used our regression framework to determine whether these LVs were significantly associated with different traits.
For example, LVs such as LV928 and LV30, which were well-aligned to early progenitors of the erythrocytes lineage\textsuperscript{\protect\hyperlink{ref-Zk82GvJV}{82}} (Supplementary Tables 12 and 15), were predominantly expressed in early differentiation stages of erythropoiesis (Supplementary Figures 17 and 18) and strongly associated with different assays on red blood cells (FDR \textless{} 0.05; Supplementary Tables 13, 14, and 17).
In contrast, other LVs were highly specific, such as LV730, which is expressed in thrombocytes from different cancer samples (Supplementary Figures 19 and Supplementary Table 18), and strongly associated with hematological assays on platelets (FDR \textless{} 0.05, Supplementary Table 19);
or LV598, whose genes were expressed in corneal endothelial cells (Supplementary Figures 20 and Supplementary Table 21) and associated with keratometry measurements (Supplementary Table 22).

The sub-branches of autoimmune and cardiovascular diseases merged together at \(k=10\) (middle of Figure \ref{fig:clustering:tree}), so we expected to find LVs that specifically affect one or both of these types of diseases.
For example, LV57, expressed in T cells (Supplementary Figure 21 and Supplementary Table 24), was the most strongly associated gene module with autoimmune disorders in PhenomeXcan (Supplementary Table 25), with significant associations with hypothyroidism that were replicated in eMERGE (Supplementary Table 26).
However, this LV was also strongly associated with deep venous thrombosis in both PhenomeXcan and eMERGE.
On the other hand, LV844 was more autoimmune-specific, with associations to polymyalgia rheumatica, type 1 diabetes, rheumatoid arthritis, and celiac disease in PhenomeXcan (Supplementary Table 28).
However, these did not replicate in eMERGE.
This LV was expressed in a wide range of cell types, including blood, breast organoids, myeloma cells, lung fibroblasts, and different cell types from the brain (Supplementary Figure 22 and Supplementary Table 27).

The cardiovascular sub-branch had 129 significant LV-trait associations in PhenomeXcan and 23 in eMERGE.
LV136, aligned with known collagen formation and muscle contraction pathways (Supplementary Table 30), was associated with coronary artery disease and keratometry measurements in PhenomeXcan (Supplementary Table 31).
In eMERGE, this LV was associated with coronary atherosclerosis (phecode: 411.4) (Supplementary Table 32).
LV136 was expressed in a wide range of cell types, including fibroblasts, mesenchymal stem cells, osteoblasts, pancreatic stellate cells, cardiomyocytes, and adipocytes (Supplementary Figure 23).
Within the cardiovascular sub-branch, we found neuropsychiatric and neurodevelopmental disorders such as Alzheimer's disease, schizophrenia, and attention deficit hyperactivity disorder (ADHD).
These disorders were previously linked to the cardiovascular system\textsuperscript{\protect\hyperlink{ref-10qjLoufR}{83}--\protect\hyperlink{ref-13t4TuFeJ}{86}} and share several risk factors, including hypertension, high cholesterol, obesity, smoking, among others\textsuperscript{\protect\hyperlink{ref-D83Aqhga}{87},\protect\hyperlink{ref-9BGyO071}{88}}.
However, our results grouped these diseases by potentially shared transcriptional processes expressed in specific tissues/cell types.
Alzheimer's disease (not present in eMERGE), for instance, was significantly associated with LV21 in PhenomeXcan (Supplementary Table 34).
LV21, a gene module not aligned to prior pathways, was strongly expressed in a variety of soft tissue sarcomas, monocytes/macrophages (including microglia from cortex samples), and aortic valves (Supplementary Figure 24 and Supplementary Table 33).
This LV was also strongly associated with lipids and high cholesterol in PhenomeXcan and hyperlipidemia (phecode: 272.1) in eMERGE (Supplementary Table 35).
As discussed previously, macrophages play a key role in the reverse cholesterol transport and thus atherogenesis\textsuperscript{\protect\hyperlink{ref-idlBgtFz}{89}}, and lipid metabolism in microglia has been recently identified as an important factor in the development of neurodegenerative diseases\textsuperscript{\protect\hyperlink{ref-18I4ish9s}{90}}.

@@

Note that, since we used the MultiXcan regression model (Equation (\ref{eq:multixcan})), \(\mathbf{R}\) is only an approximation of gene correlations in S-MultiXcan.
As explained before, S-MultiXcan approximates the joint regression parameters in MultiXcan using the marginal regression estimates from S-PrediXcan in (\ref{eq:spredixcan}) with some simplifying assumptions and different genotype covariance matrices.
This complicates the derivation of an S-MultiXcan-specific solution to compute \(\mathbf{R}\).
To account for this, we used a submatrix \(\mathbf{R}_{\ell}\) corresponding to genes that are part of LV \(\ell\) only (top 1\% of genes) instead of the entire matrix \(\mathbf{R}\).
This simplification is conservative since correlations are accounted for top genes only.
Our simulations (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) show that the model is approximately well-calibrated and can correct for LVs with adjacent and highly correlated genes at the top (e.g., Supplementary Figure 2).
The simulation also identified 127 LVs in which the model was not well-calibrated (e.g., Supplementary Figure 6).
As this can be attributed to limitations in accurately computing a gene correlation matrix, we excluded these LVs from our main analyses.

@@

Finally, we used spectral clustering on \(\mathbf{D}\) to derive the final consensus partitions.
\(\mathbf{D}\) was first transformed into a similarity matrix by applying an RBF kernel \(\mathrm{exp}(-\gamma \mathbf{D}^2)\) using four different values for \(\gamma\) that we empirically determined to work best.
Therefore, for each \(k\) between 2 and 60, we derived four consensus partitions and selected the one that maximized Equation (\ref{eq:consensus:obj_func}).
We further filtered this set of 59 solutions to keep only those with an ensemble agreement larger than the 75th percentile (Supplementary Figure 13), leaving a total of 15 final consensus partitions shown in Figure \ref{fig:clustering:tree}.
